# Data

## Context & Background Information

This dataset originates from the National Institute of Diabetes and Digestive and Kidney Diseases.
Several constraints were applied when selecting instances from a larger database. In particular, all patients are females at least 21 years old and of Pima Indian heritage. The dataset consists of 768 samples, with 8 variables and an outcome column. The dataset was obtained from Kaggle.

## Content

- Number of Sample: 768

- Number of Attribute: 8

- Pregnancies: number of times pregnant
- Glucose: plasma glucose concentration 2 hours in an oral glucose tolerance
test
- BloodPressure: diastolic blood pressure (mm Hg)
- SkinThinkness: triceps skin fold thickness (mm)
- Insulin: 2-Hour serum insulin (mu U/ml)
- BMI: body mass index (weight in kg/(height in m)^2)
- DiabetesPedigreeFunction: diabetes pedigree function
- Age: age (years)
- Outcome: class variable (0 = No, 1 = Yes)


```{r}
library(dplyr)
df <- read.csv("data/diabetes.csv")
summary(df)
# Summary of 0
sapply(df, function(x) sum(x == 0))
```

## Missing value analysis
This dataset contains missing values, with ‘0’ used to fill them beforehand. As a result, it is difficult to determine whether a ‘0’ represents a missing value or an actual measurement of zero.

From the summary, variable 'Pregnancies' have 111 responses are '0' which I will not mutate this column. Since this dataset includes only females who are at least 21 years old, I believe that some of the zeros in the pregnancies variable likely indicate no pregnancy history.

There is 5 missing values for 'Glucose', 35 missing values for 'BloodPressure', 277 missing values for 'SkinThickness', 374 missing values for 'Insulin', 11 missing values for 'BMI', and no missing values for 'DiabetesPedigreeFunction' and 'Age'. I would replace zeros with the median of the non-zero values. The median is robust to outliers and is commonly used in medical datasets.

```{r}
missing_cols <- c("Glucose", "BloodPressure", "SkinThickness", "Insulin", "BMI")

# Replace 0 with median
for (col in missing_cols) {
  median_value <- median(df[[col]][df[[col]] != 0], na.rm = TRUE)
  df[[col]][df[[col]] == 0] <- median_value
}

sapply(df, function(x) sum(x == 0))
```

In my study, the variable I propose to categorize and make inferences about "Outcome", indicates the potential diagnosis of diabetes. This variable determines whether a diabetes test result is positive (category value 1) or negative (category value 0). One notable aspect of our dataset is the “Pregnancies” variable, which has a minimum value of 0, a maximum value of 17, and a mean of 3.845. It would be particularly informative to explore whether pregnancy is an important factor influencing the outcome of a diabetes diagnosis and to assess the degree to which it does so. Additionally, common predictors such as body mass index (BMI) and age are also expected to play important roles in determining outcomes.


